---
title: "USING FGconnectionHandler: Adding names to longitudinal and service sector codes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial_add_info_FGconnectionHandler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Intro

This tutorial show how to append additional information to a longitudinal or service-sector table using `dplyr` package and the `FGconnectionHandler` object in the `FinnGenUtilsR` package. 

Information to add includes, name of the medical codes in English, name for the type of visit, name for the type of provider.

To understand what is the `FGconnectionHandler` is recommended to see the dedicate vignette `tutorial_connection_handler`.


# Set up

```{r setup}
library(dplyr)
#library(FinnGenUtilsR)
devtools::load_all(".")
```

This tutorial has been developed using dummy data under the Atlas development project. 

```{r}
 config <- fg_get_bq_config(
   environment = "atlasDevelopment",
   dataFreezeNumber = 11
   )
```


To execute it in Sandbox, be sure to change the configuration to fit your needs. 
```{r, eval=FALSE}
 config <- fg_get_bq_config(
   environment = "sandbox-6", # set your sandbox number
   dataFreezeNumber = 12 # set the dataFreeze you want to use
   )
```

Create a `FGconnectionHandler` object with the connection configuration from `fg_get_bq_config`. 

```{r}
FGconnectionHandler <- create_fg_connection_handler_FromList(config)
```

We can check if the connection has been successful. 

```{r}
FGconnectionHandler$connectionStatusLog
```

We can access the service sector data using the connection object as follows: 

```{r}
service_sector_tbl <- FGconnectionHandler$getTblsandboxToolsSchema$finngen_r11_service_sector_detailed_longitudinal_v1()
```

Where `service_sector_tbl` is a `tbl` object that can be used with `dplyr` functions. 

# Generate a table in service sector format

For example we create a subset of the main service sector table by filtering the patients with an asthma diagnose (J45) in the INPAT source. 


**IMPORTANT:** this is important to notice that the tables coming from the `FGconnectionHandler` have all the columns in lower case.

```{r}
# get events with code "J45" in INPAT
ss_subject_1_tbl  <- FGconnectionHandler$getTblsandboxToolsSchema$finngen_r11_service_sector_detailed_longitudinal_v1() |> 
  dplyr::filter(finngenid == "FG00000001")
```

We can see that this table does not currently have information about the medical codes: 

```{r}
ss_subject_1_tbl
```
(same table in html format for exporation)
```{r echo=FALSE}
ss_subject_1_tbl |> as.data.frame() |> rmarkdown::paged_table()
```

# Adding info to medical codes

Function `fg_dbplyr_append_code_info_to_longitudinal_data` adds new columns with information about the medical code used in each event/row.  

We just need to indicate the table with the traslations. This is also available in the `FGconnectionHandler` object. 

We recommend to take the one with the hihgset version number. `fg_codes_info_v6` on the moment of writing.


```{r}
ss_subject_1_with_translations_tbl <- fg_dbplyr_append_code_info_to_longitudinal_data(
  ss_subject_1_tbl, 
  FGconnectionHandler$getTblmedicalCodesSchema$fg_codes_info_v6()
  )
```
 
```{r}
ss_subject_1_with_translations_tbl |>  
  dplyr::select(finngenid, source, approx_event_day, code1, code2, code3, code, name_en, name_fi, omop_concept_id) 
```
(same table in html format for exporation)
```{r echo=FALSE}
ss_subject_1_with_translations_tbl |>  
  dplyr::select(finngenid, source, approx_event_day, code1, code2, code3, code, name_en, name_fi, omop_concept_id)  |> 
  as.data.frame() |> rmarkdown::paged_table()
```

## If an event/row uses more than one medical code 

Some events include more than one medical code, or more than one way to understand a medical code. 
Parameters in `fg_dbplyr_append_code_info_to_longitudinal_data` allows to chose how to name the medical code in this situations. 

For example: 

 - PURCH_map_to = "VNR", tells to use the VNR code in PURCH register, instead of the default ATC
 - ICD10fi_map_to = "CODE1", tells to use only the first code in ICD10fi, instead of the default CODE1*CODE2 combination code
 
```{r}
fg_dbplyr_append_code_info_to_longitudinal_data(
  ss_subject_1_tbl, 
  FGconnectionHandler$getTblmedicalCodesSchema$fg_codes_info_v6(), 
    PURCH_map_to = "VNR", 
    ICD10fi_map_to = "CODE1" 
  ) |>
  filter(vocabulary_id == "ICD10fi" | vocabulary_id == "VNRfi") |>  
  dplyr::select(finngenid, source, approx_event_day, code1, code2, vocabulary_id, code, name_en, name_fi, omop_concept_id) 
```

(same table in html format for exporation)
```{r echo=FALSE}
ss_subject_1_with_translations_tbl |>  
  dplyr::select(finngenid, source, approx_event_day, code1, code2, code3, code, name_en, name_fi, omop_concept_id)  |> 
  as.data.frame() |> rmarkdown::paged_table()
```

Se the help, for more options

```{r, eval=FALSE}
?fg_append_code_info_to_longitudinal_data_sql
```


## Reduce code precision

ICD10fi, ICD9fi, ICD8fi, ATC, and NCSPfi, code systems reflect the hierarchy of the code by the first letters on them. 

It is possible to truncate these medical codes before the information is added. This in practice adds the code and names of the parent codes. 

```{r}
fg_dbplyr_append_code_info_to_longitudinal_data(
  ss_subject_1_tbl, 
  FGconnectionHandler$getTblmedicalCodesSchema$fg_codes_info_v6(), 
  ICD10fi_precision = 3,
  ICD9fi_precision = 3,
  ICD8fi_precision = 3,
  ATC_precision = 3,
  NCSPfi_precision = 2
)|>
  dplyr::select(finngenid, source, approx_event_day, code1, code2, vocabulary_id, code, name_en, name_fi, omop_concept_id) 

```
(same table in html format for exporation)
```{r echo=FALSE}
ss_subject_1_with_translations_tbl |>  
  dplyr::select(finngenid, source, approx_event_day, code1, code2, code3, code, name_en, name_fi, omop_concept_id)  |> 
  as.data.frame() |> rmarkdown::paged_table()
```

We can see in column `code` the truncated code and in `name_en` the name of the truncated code. 


# Adding info to visit type 

Function `fg_dbplyr_append_visit_type_info_to_service_sector_data` adds new columns with information about the visit type.

In the service-sector data, information about the visit type is defined in CODE5 to CODE9 depending on the SOURCE and time period. 
`fg_dbplyr_append_visit_type_info_to_service_sector_data` abstracts this nuances, and assign one visit type code per row. 
Notice that a visit may contain more than one event/row. Events/rows belonging to the same visit share the same combination of SOURCE+INDEX. 


```{r}
fg_dbplyr_append_visit_type_info_to_service_sector_data(
  ss_subject_1_tbl, 
  FGconnectionHandler$getTblmedicalCodesSchema$fg_codes_info_v6()
  )|> 
  dplyr::select(finngenid, source, index, approx_event_day, code5, code6, code8, code9, visit_type_code, visit_type_name_en)
```

(same table in html format for exporation)
```{r echo=FALSE}
ss_subject_1_with_translations_tbl |>  
  dplyr::select(finngenid, source, approx_event_day, code1, code2, code3, code, name_en, name_fi, omop_concept_id)  |> 
  as.data.frame() |> rmarkdown::paged_table()
```

# Adding info to provider type 

Function `fg_dbplyr_append_provider_info_to_service_sector_data` adds new columns with information about the personal or unit that provided the diagnose during the visit.

In the service-sector data, information about the provider is defined in CODE6 to CODE7 depending on the SOURCE. 
`fg_dbplyr_append_provider_info_to_service_sector_data` abstracts this nuances, and assign one provider type code per row. 


```{r}
fg_dbplyr_append_provider_info_to_service_sector_data(
  ss_subject_1_tbl, 
  FGconnectionHandler$getTblmedicalCodesSchema$fg_codes_info_v6()
  )|> 
  dplyr::select(finngenid, source, index, approx_event_day, code5, code6, code8, code9, provider_code, provider_name_en, provider_concept_class_id)
```
(same table in html format for exporation)
```{r echo=FALSE}
fg_dbplyr_append_provider_info_to_service_sector_data(
  ss_subject_1_tbl, 
  FGconnectionHandler$getTblmedicalCodesSchema$fg_codes_info_v6()
  )|> 
  dplyr::select(finngenid, source, index, approx_event_day, code5, code6, code8, code9, provider_code, provider_name_en, provider_concept_class_id) |> 
  as.data.frame() |> rmarkdown::paged_table()
```










