---
title: "Working with Kanta Lab Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial_kanta_lab_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Intro

This tutorial demonstrates how to work with Kanta laboratory data using the `FinnGenUtilsR` package. The Kanta lab data contains laboratory measurements with OMOP concept mappings, and this tutorial shows how to append human-readable names to these OMOP concept IDs.

To understand what the `fgbq` object is, see the dedicated vignette `tutorial_fgbq`.

# Set up

```{r setup}
library(dplyr)
library(FinnGenUtilsR)
```

Connect to the latest versions of BigQuery tables using `get_fg_bq_tables()`.
Remember to change the environment to your sandbox, or set other data freeze versions if needed (see `tutorial_fgbq` vignette for more details).

```{r}
fgbq <- get_fg_bq_tables(
  environment = "sandbox-XX"
)
```

We can access the Kanta lab data using the `fgbq` object as follows: 

```{r}
fgbq$tbl$kanta
```

# Generate a table for analysis

For example we create a subset of the main Kanta table by querying latest events from subject "FG00000001": 

```{r}
kanta_subject_1_tbl <- fgbq$tbl$kanta |> 
  dplyr::filter(FINNGENID == "FG00000001") |> 
  dplyr::arrange(desc(APPROX_EVENT_DATETIME)) 
```

We can see that this table contains OMOP_CONCEPT_ID but not the names of the concepts: 

```{r}
kanta_subject_1_tbl
```
(same table in html format for exploration)
```{r echo=FALSE}
kanta_subject_1_tbl |> as.data.frame() |> rmarkdown::paged_table()
```


## Adding info to OMOP_CONCEPT_ID codes

Function `fg_dbplyr_append_concept_info_data` adds a new column with information about the OMOP_CONCEPT_ID name.

```{r}
omop_schema <- "atlas-development-270609.etl_sam_r12_5k_omop"
kanta_subject_1_tbl_with_omop_name <- kanta_subject_1_tbl |> fg_dbplyr_append_concept_info_data(omop_schema)
```


```{r}
kanta_subject_1_tbl_with_omop_name |>  
  dplyr::select(FINNGENID, EVENT_AGE, APPROX_EVENT_DATETIME, TEST_NAME, TEST_ID, OMOP_CONCEPT_ID, omop_concept_name) 
```
(same table in html format for exploration)
```{r echo=FALSE}
kanta_subject_1_tbl_with_omop_name |>  
  dplyr::select(FINNGENID, EVENT_AGE, APPROX_EVENT_DATETIME, TEST_NAME, TEST_ID, OMOP_CONCEPT_ID, omop_concept_name) |> 
  dplyr::collect() |> 
  rmarkdown::paged_table()
```
