---
title: "Adding names to longitudinal and service sector codes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial_add_info_fgbq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Intro

This tutorial shows how to append additional information to a longitudinal or service-sector table using `dplyr` package and the `fgbq` object from the `FinnGenUtilsR` package. 

Information to add includes: name of the medical codes in English, name for the type of visit, name for the type of provider.

To understand what the `fgbq` object is, see the dedicated vignette `tutorial_fgbq`.

This tutorial also includes how to append additional information to the Kanta lab data. In particular, the name of the OMOP concept for the lab test.

# Set up

```{r setup}
library(dplyr)
library(FinnGenUtilsR)
```

Connect to the latest versions of BigQuery tables using `get_fg_bq_tables()`.
Remember to change the environment to your sandbox, or set other data freeze versions if needed (see `tutorial_fgbq` vignette for more details).

```{r}
fgbq <- get_fg_bq_tables(
  environment = "sandbox-XX"
)
```

We can access the service sector data using the `fgbq` object as follows: 

```{r}
fgbq$tbl$service_sector_detailed_longitudinal
```

# Generate a table in service sector format

For example we create a subset of the main service sector table by filtering events for one patient.

```{r}
# Get events for subject FG00000001
ss_subject_1 <- fgbq$tbl$service_sector_detailed_longitudinal |> 
  dplyr::filter(FINNGENID == "FG00000001")
```

We can see that this table does not currently have information about the medical codes: 

```{r}
ss_subject_1
```
(same table in html format for exporation)
```{r echo=FALSE}
ss_subject_1_tbl |> as.data.frame() |> rmarkdown::paged_table()
```

## Adding info to medical codes

Function `fg_dbplyr_append_code_info_to_longitudinal_data` adds new columns with information about the medical code used in each event/row.  

We just need to indicate the table with the translations. This is also available in the `fgbq` object.

```{r}
ss_subject_1_with_translations_tbl <- fg_dbplyr_append_code_info_to_longitudinal_data(
  ss_subject_1_tbl, 
  fgbq$tbl$fg_codes_info
)
```
 
```{r}
ss_subject_1_with_translations_tbl |>  
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, CODE3, code, name_en, name_fi, omop_concept_id) 
```
(same table in html format for exploration)
```{r echo=FALSE}
ss_subject_1_with_translations_tbl |>  
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, CODE3, code, name_en, name_fi, omop_concept_id) |> 
  as.data.frame() |> rmarkdown::paged_table()
```

## If an event/row uses more than one medical code 

Some events include more than one medical code, or more than one way to understand a medical code. 
Parameters in `fg_dbplyr_append_code_info_to_longitudinal_data` allows to chose how to name the medical code in this situations. 

For example: 

 - PURCH_map_to = "VNR", tells to use the VNR code in PURCH register, instead of the default ATC
 - ICD10fi_map_to = "CODE1", tells to use only the first code in ICD10fi, instead of the default CODE1*CODE2 combination code
 
```{r}
fg_dbplyr_append_code_info_to_longitudinal_data(
  ss_subject_1_tbl, 
  fgbq$tbl$fg_codes_info, 
  PURCH_map_to = "VNR", 
  ICD10fi_map_to = "CODE1" 
) |>
  filter(vocabulary_id == "ICD10fi" | vocabulary_id == "VNRfi") |>  
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, vocabulary_id, code, name_en, name_fi, omop_concept_id) 
```

(same table in html format for exploration)
```{r echo=FALSE}
ss_subject_1_with_translations_tbl |>  
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, CODE3, code, name_en, name_fi, omop_concept_id) |> 
  as.data.frame() |> rmarkdown::paged_table()
```

Se the help, for more options

```{r, eval=FALSE}
?fg_append_code_info_to_longitudinal_data_sql
```


## Reduce code precision

ICD10fi, ICD9fi, ICD8fi, ATC, and NCSPfi, code systems reflect the hierarchy of the code by the first letters on them. 

It is possible to truncate these medical codes before the information is added. This in practice adds the code and names of the parent codes. 

```{r}
fg_dbplyr_append_code_info_to_longitudinal_data(
  ss_subject_1_tbl, 
  fgbq$tbl$fg_codes_info, 
  ICD10fi_precision = 3,
  ICD9fi_precision = 3,
  ICD8fi_precision = 3,
  ATC_precision = 3,
  NCSPfi_precision = 2
) |>
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, vocabulary_id, code, name_en, name_fi, omop_concept_id) 

```
(same table in html format for exploration)
```{r echo=FALSE}
ss_subject_1_with_translations_tbl |>  
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, CODE3, code, name_en, name_fi, omop_concept_id) |> 
  as.data.frame() |> rmarkdown::paged_table()
```

We can see in column `code` the truncated code and in `name_en` the name of the truncated code. 


# Adding info to visit type 

Function `fg_dbplyr_append_visit_type_info_to_service_sector_data` adds new columns with information about the visit type.

In the service-sector data, information about the visit type is defined in CODE5 to CODE9 depending on the SOURCE and time period. 
`fg_dbplyr_append_visit_type_info_to_service_sector_data` abstracts this nuances, and assign one visit type code per row. 
Notice that a visit may contain more than one event/row. Events/rows belonging to the same visit share the same combination of SOURCE+INDEX. 


By default, `fg_bq_append_visit_type_info_to_service_sector_data` will also include two columns `is_clinic_visit` and `is_follow_up_visit` that will be TRUE if the visit is a clinic visit or a follow-up visit, respectively. However, this works only if `fg_codes_info_table` version is v7 or higher. If you need to use lower versions set the parameters `add_is_clinic_visist` and `add_is_follow_up_visit` to FALSE. See the help for more details.

```{r}
fg_dbplyr_append_visit_type_info_to_service_sector_data(
  ss_subject_1_tbl, 
  fgbq$tbl$fg_codes_info
) |> 
  dplyr::select(FINNGENID, SOURCE, INDEX, APPROX_EVENT_DAY, CODE5, CODE6, CODE8, CODE9, visit_type_code, visit_type_name_en, is_clinic_visit, is_follow_up_visit)
```

(same table in html format for exploration)
```{r echo=FALSE}
fg_dbplyr_append_visit_type_info_to_service_sector_data(
  ss_subject_1_tbl, 
  fgbq$tbl$fg_codes_info
) |> 
  dplyr::select(FINNGENID, SOURCE, INDEX, APPROX_EVENT_DAY, CODE5, CODE6, CODE8, CODE9, visit_type_code, visit_type_name_en, is_clinic_visit, is_follow_up_visit) |> 
  as.data.frame() |> rmarkdown::paged_table()
```

# Adding info to provider type 

Function `fg_dbplyr_append_provider_info_to_service_sector_data` adds new columns with information about the personal or unit that provided the diagnose during the visit.

In the service-sector data, information about the provider is defined in CODE6 to CODE7 depending on the SOURCE. 
`fg_dbplyr_append_provider_info_to_service_sector_data` abstracts this nuances, and assign one provider type code per row. 


```{r}
fg_dbplyr_append_provider_info_to_service_sector_data(
  ss_subject_1_tbl, 
  fgbq$tbl$fg_codes_info
) |> 
  dplyr::select(FINNGENID, SOURCE, INDEX, APPROX_EVENT_DAY, CODE5, CODE6, CODE8, CODE9, provider_code, provider_name_en, provider_concept_class_id)
```
(same table in html format for exploration)
```{r echo=FALSE}
fg_dbplyr_append_provider_info_to_service_sector_data(
  ss_subject_1_tbl, 
  fgbq$tbl$fg_codes_info
) |> 
  dplyr::select(FINNGENID, SOURCE, INDEX, APPROX_EVENT_DAY, CODE5, CODE6, CODE8, CODE9, provider_code, provider_name_en, provider_concept_class_id) |> 
  as.data.frame() |> rmarkdown::paged_table()
```



# Alternative: Running SQL Queries

The previous examples used `dplyr` syntax with `fgbq$tbl` objects. Alternatively, you can work directly with BigQuery tables using the `fg_bq_*` family of functions, which execute SQL queries directly on BigQuery.

This approach is useful when you need more control over the query execution or when working with existing BigQuery table references.

## Setup for BigQuery approach

First, we need to set up the BigQuery connection and table references:

```{r}
library(bigrquery)

# Your BigQuery project ID
project_id <- "atlas-development-270609"

# Path to the fg_codes_info table
fg_codes_info_table <- "atlas-development-270609.medical_codes_dev.fg_codes_info_dev"

# Create a bq_table reference to your service sector data
# This could be a table you've already created in BigQuery
service_sector_bq_table <- bq_table(
  project = project_id,
  dataset = "sandbox_tools_dev",
  table = "finngen_dev_service_sector_detailed_longitudinal_dev"
)
```

## Generate a table in service sector format

```{r}
# Get events for subject FG00000001 directly from BigQuery
ss_subject_1_bq <- bq_project_query(
  project_id,
  "SELECT * FROM `atlas-development-270609.sandbox_tools_dev.finngen_dev_service_sector_detailed_longitudinal_dev` WHERE FINNGENID = 'FG00000001'"
) 
```

## Adding info to medical codes with BigQuery

Function `fg_bq_append_code_info_to_longitudinal_data` works similarly to the `dplyr` version, but operates directly on BigQuery tables:

```{r}
# Add code information to the service sector table
result_bq_table <- fg_bq_append_code_info_to_longitudinal_data(
  bq_project_id = project_id,
  bq_table = ss_subject_1_bq,
  fg_codes_info_table = fg_codes_info_table
)

# The result is a new temporary table in BigQuery
result_bq_table |> bq_table_download() |>  
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, vocabulary_id, code, name_en, name_fi, omop_concept_id)
```
