---
title: "USING BIGRQUERY PACKAGE: Adding names to longitudinal and service sector codes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial_add_info_bigrquery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Intro

This tutorial show how to append additional information to a longitudinal or service-sector table in bigquery using the `bigrquery` package. 

Information to add includes, name of the medical codes in English, name for the type of visit, name for the type of provider.


# Set up

This tutorial has been developed using dummy data under the Atlas development project. 
To execute it in Sandbox, be sure to change the following paths. 

```{r setup}
library(dplyr)
library(bigrquery)
library(FinnGenUtilsR)


bigrquery::bq_auth(path = Sys.getenv("GCP_SERVICE_KEY"))
project_id <- "atlas-development-270609"

test_longitudinal_data_table <- "atlas-development-270609.sandbox_tools_r10.finngen_r10_service_sector_detailed_longitudinal_v2"
fg_codes_info_table <- "atlas-development-270609.medical_codes.fg_codes_info_v2"
vocab_omop_concept_relationship <- "atlas-development-270609.finngen_omop_r11.concept_relationship"
vocab_omop_concept <- "atlas-development-270609.finngen_omop_r11.concept"
tmp_schema <- "sandbox"
```


# Generate a table in service sector format

For example we create a subset of the main service sector table by querying latest 20 events from subject "FG00000001": 

```{r}
sql <- paste("SELECT * FROM ", test_longitudinal_data_table, "WHERE FINNGENID='FG00000001' ORDER BY APPROX_EVENT_DAY DESC LIMIT 20")
event_for_subject_1 <- bq_project_query(project_id, sql)
```

We can see that this table does not currently have information about the medical codes: 

```{r}
bq_table_download(event_for_subject_1)
```
(same table in html format for exporation)
```{r echo=FALSE}
bq_table_download(event_for_subject_1) |> rmarkdown::paged_table()
```

# Adding info to medical codes

Function `fg_bq_append_code_info_to_longitudinal_data` adds new columns with information about the medical code used in each event/row.  

```{r}
event_for_subject_1_with_translations <- fg_bq_append_code_info_to_longitudinal_data(
  project_id, event_for_subject_1, fg_codes_info_table)
```

For example: code, for the original code; name_en, for the name of the code in English; name_fi for the name of the code in Finnish; and omop_concept_id, to find additional info in Atlas. 

```{r}
bq_table_download(event_for_subject_1_with_translations) |>  
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, CODE3, code, name_en, name_fi, omop_concept_id) 
```
(same table in html format for exporation)
```{r echo=FALSE}
bq_table_download(event_for_subject_1_with_translations) |>  
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, CODE3, code, name_en, name_fi, omop_concept_id)  |>
  rmarkdown::paged_table()
```

## If an event/row uses more than one medical code 

Some events include more than one medical code, or more than one way to understand a medical code. 
Parameters in `fg_bq_append_code_info_to_longitudinal_data` allows to chose how to name the medical code in this situations. 

For example: 

 - PURCH_map_to = "VNR", tells to use the VNR code in PURCH register, instead of the default ATC
 - ICD10fi_map_to = "CODE1", tells to use only the first code in ICD10fi, instead of the default CODE1*CODE2 combination code
 
```{r}
event_for_subject_1_with_translations <- fg_bq_append_code_info_to_longitudinal_data(
  project_id, event_for_subject_1, fg_codes_info_table, 
    PURCH_map_to = "VNR", 
    ICD10fi_map_to = "CODE1"
)
one_patient_long_data <- bq_table_download(event_for_subject_1_with_translations)
```

```{r}
bq_table_download(event_for_subject_1_with_translations) |>  
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, CODE3, code, name_en, name_fi, omop_concept_id) 
```

(same table in html format for exporation)
```{r echo=FALSE}
bq_table_download(event_for_subject_1_with_translations) |>  
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, CODE3, code, name_en, name_fi, omop_concept_id)  |> 
  rmarkdown::paged_table()
```

Se the help, for more options

```{r, eval=FALSE}
?fg_append_code_info_to_longitudinal_data_sql
```


## Reduce code precision

ICD10fi, ICD9fi, ICD8fi, ATC, and NCSPfi, code systems reflect the hierarchy of the code by the first letters on them. 

It is possible to truncate these medical codes before the information is added. This in practice adds the code and names of the parent codes. 

```{r}
event_for_subject_1_with_translations <- fg_bq_append_code_info_to_longitudinal_data(
  project_id, event_for_subject_1, fg_codes_info_table, 
  ICD10fi_precision = 3,
  ICD9fi_precision = 3,
  ICD8fi_precision = 3,
  ATC_precision = 3,
  NCSPfi_precision = 2
)

```

We can see in column `code` the trucated code and in `name_en` the name of the trucated code. 

```{r}
bq_table_download(event_for_subject_1_with_translations) |>  
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, CODE3, code, name_en, name_fi, omop_concept_id) 
```




(same table in html format for exporation)
```{r echo=FALSE}
bq_table_download(event_for_subject_1_with_translations) |>  
  dplyr::select(FINNGENID, SOURCE, APPROX_EVENT_DAY, CODE1, CODE2, CODE3, code, name_en, name_fi, omop_concept_id)  |>
  rmarkdown::paged_table()
```



# Adding info to visit type 

Function `fg_bq_append_visit_type_info_to_service_sector_data` adds new columns with information about the visit type.

In the service-sector data, information about the visit type is defined in CODE5 to CODE9 depending on the SOURCE and time period. 
`fg_bq_append_visit_type_info_to_service_sector_data` abstracts this nuances, and assign one visit type code per row. 
Notice that a visit may contain more than one event/row. Events/rows belonging to the same visit share the same combination of SOURCE+INDEX. 


```{r}
event_for_subject_1_with_visit_type <- fg_bq_append_visit_type_info_to_service_sector_data(
  project_id, event_for_subject_1, fg_codes_info_table, vocab_omop_concept_relationship, vocab_omop_concept
  )
```


```{r}
bigrquery::bq_table_download(event_for_subject_1_with_visit_type) |> 
  dplyr::select(FINNGENID, SOURCE, INDEX, APPROX_EVENT_DAY, CODE5, CODE6, CODE8, CODE9, visit_type_code, visit_type_name_en, diagnostic_relevance)
```


(same table in html format for exporation)
```{r echo=FALSE}
bigrquery::bq_table_download(event_for_subject_1_with_visit_type) |> 
  dplyr::select(FINNGENID, SOURCE, INDEX, APPROX_EVENT_DAY, CODE5, CODE6, CODE8, CODE9, visit_type_code, visit_type_name_en, diagnostic_relevance)|> 
  rmarkdown::paged_table()
```



# Adding info to provider type 

Function `fg_bq_append_provider_info_to_service_sector_data` adds new columns with information about the personal or unit that provided the diagnose during the visit.

In the service-sector data, information about the provider is defined in CODE6 to CODE7 depending on the SOURCE. 
`fg_bq_append_visit_type_info_to_service_sector_data` abstracts this nuances, and assign one provider type code per row. 


```{r}
event_for_subject_1_with_provider <- fg_bq_append_provider_info_to_service_sector_data(
  project_id, event_for_subject_1, fg_codes_info_table
  )
```


```{r}
bigrquery::bq_table_download(event_for_subject_1_with_provider) |> 
  dplyr::select(FINNGENID, SOURCE, INDEX, APPROX_EVENT_DAY, CODE5, CODE6, CODE8, CODE9, provider_code, provider_name_en, provider_concept_class_id)
```



(same table in html format for exporation)
```{r echo=FALSE}
bigrquery::bq_table_download(event_for_subject_1_with_provider) |> 
  dplyr::select(FINNGENID, SOURCE, INDEX, APPROX_EVENT_DAY, CODE5, CODE6, CODE8, CODE9, provider_code, provider_name_en, provider_concept_class_id) |> rmarkdown::paged_table()
```







