---
title: "Adding Names to Drug Events Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial_drug_events_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Intro

This tutorial demonstrates how to work with drug events data using the `FinnGenUtilsR` package. The drug events data contains medication purchase and prescription records with VNR (Finnish national product codes) identifiers. This tutorial shows how to append human-readable medication names and ATC codes by joining with the `finngen_vnrs` reference table.

Unlike other data types (such as service sector or longitudinal data), there is no custom helper function for adding names to drug events data. Instead, we use a simple join operation with the `finngen_vnrs` table.

To understand what the `fgbq` object is, see the dedicated vignette `tutorial_fgbq`.

# Set up

```{r setup}
library(dplyr)
library(FinnGenUtilsR)
```

Connect to the latest versions of BigQuery tables using `get_fg_bq_tables()`.
Remember to change the environment to your sandbox, or set other data freeze versions if needed (see `tutorial_fgbq` vignette for more details).

```{r}
fgbq <- get_fg_bq_tables(
  environment = "sandbox-XX"
)
```

We can access the drug events data using the `fgbq` object as follows: 

```{r}
fgbq$tbl$drug_events |> 
  glimpse()
```

We can also view the finngen_vnrs reference table that contains medication names and ATC codes:

```{r}
fgbq$tbl$finngen_vnrs |> 
  glimpse()
```

# Generate a table for analysis

For example, we create a subset of the drug events table by querying events from subject "FG00003640": 

```{r}
drug_events_subject_tbl <- fgbq$tbl$drug_events |> 
  filter(FINNGENID == "FG00003640") |> 
  arrange(desc(MEDICATION_APPROX_EVENT_DAY))
```

We can see that this table contains VNR codes (MEDICATION_VNR and PRESCRIPTION_VNR) but not the medication names:

```{r}
drug_events_subject_tbl |> 
  glimpse()
``` 


## Adding medication information via VNR codes

To add medication names and ATC codes to the drug events data, we simply join with the `finngen_vnrs` reference table. The drug events table contains two VNR columns:

- `MEDICATION_VNR`: The VNR code for the actual medication dispensed
- `PRESCRIPTION_VNR`: The VNR code from the original prescription (may differ from what was dispensed)

We can join on either column depending on which medication information we need. Here we'll join on `MEDICATION_VNR`:

Note: The VNR codes in the drug_events table are stored as strings, while in the finngen_vnrs table they are integers. We need to convert the type before joining.

```{r}
drug_events_with_names <- drug_events_subject_tbl |> 
  mutate(MEDICATION_VNR = as.integer(MEDICATION_VNR)) |> 
  left_join(
    fgbq$tbl$finngen_vnrs,
    by = c("MEDICATION_VNR" = "VNR")
  )
```

Now we can see the medication names, ATC codes, and other information:

```{r}
drug_events_with_names |> 
  dplyr::glimpse()
```

# Alternative: Running SQL Queries
You can also run SQL queries directly on the BigQuery tables. See the `tutorial_service_sector_data` vignette for examples of how to do this with the service sector data, but the same principles apply to the drug events data.