---
title: "Adding Names to Kanta Lab Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial_kanta_lab_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Intro

This tutorial demonstrates how to work with Kanta laboratory data using the `FinnGenUtilsR` package. The Kanta lab data contains laboratory measurements with OMOP concept mappings, and this tutorial shows how to append human-readable names to these OMOP concept IDs.

To understand what the `fgbq` object is, see the dedicated vignette `tutorial_fgbq`.

# Set up

```{r setup}
library(dplyr)
library(FinnGenUtilsR)
```

Connect to the latest versions of BigQuery tables using `get_fg_bq_tables()`.
Remember to change the environment to your sandbox, or set other data freeze versions if needed (see `tutorial_fgbq` vignette for more details).

```{r}
fgbq <- get_fg_bq_tables(
  environment = "sandbox-XX"
)
```

We can access the Kanta lab data using the `fgbq` object as follows: 

```{r}
fgbq$tbl$kanta |> 
  glimmpse()
```

# Generate a table for analysis

For example we create a subset of the main Kanta table by querying latest events from subject "FG00000001": 

```{r}
kanta_subject_1_tbl <- fgbq$tbl$kanta |> 
  filter(FINNGENID == "FG00000001") |> 
  arrange(desc(APPROX_EVENT_DATETIME)) 
```

We can see that this table contains OMOP_CONCEPT_ID but not the names of the concepts: 

```{r}
kanta_subject_1_tbl |> 
  select(FINNGENID, EVENT_AGE, APPROX_EVENT_DATETIME, TEST_NAME, TEST_ID, OMOP_CONCEPT_ID) |> 
  glimmpse()
```

## Adding info to OMOP_CONCEPT_ID codes

Function `fg_dbplyr_append_code_info_to_kanta_data` adds a new column with information about the OMOP_CONCEPT_ID name.

```{r}
kanta_subject_1_tbl_with_omop_name <- kanta_subject_1_tbl |> 
   fg_dbplyr_append_code_info_to_kanta_data(fg_bq_tables = fgbq)
```


```{r}
kanta_subject_1_tbl_with_omop_name |>  
  dplyr::select(FINNGENID, EVENT_AGE, APPROX_EVENT_DATETIME, TEST_NAME, TEST_ID, OMOP_CONCEPT_ID, concept_name) |> 
  glimmpse()
```


# Alternative: Running SQL Queries
You can also run SQL queries directly on the BigQuery tables. See the `tutorial_service_sector_data` vignette for examples of how to do this with the service sector data, but the same principles apply to the Kanta lab data.